% Universal Workflow System - PROMISE 2026 Version
% Target Venue: PROMISE 2026 (Predictive Models and Data Analytics in SE)
% Format: ACM sigconf (FSE Companion Proceedings)

\documentclass[acmsmall,screen,review,anonymous]{acmart}

% ============================================================================
% PACKAGES
% ============================================================================
\usepackage{booktabs}       % Professional tables
\usepackage{subcaption}     % Subfigures
\usepackage{listings}       % Code listings
\usepackage{xcolor}         % Colors
\usepackage{tikz}           % Diagrams
\usepackage{pgfplots}       % Plots
\usepackage{algorithm}      % Algorithms
\usepackage{algorithmic}    % Algorithm formatting
\usepackage{balance}        % Balance columns on last page

\pgfplotsset{compat=1.18}

% ============================================================================
% LISTINGS CONFIGURATION
% ============================================================================
\lstset{
    basicstyle=\ttfamily\small,
    breaklines=true,
    frame=single,
    numbers=left,
    numberstyle=\tiny\color{gray},
    keywordstyle=\color{blue},
    commentstyle=\color{green!50!black},
    stringstyle=\color{red},
    showstringspaces=false,
    tabsize=2
}

% Define bash language for listings
\lstdefinelanguage{bash}{
    keywords={if, then, else, elif, fi, for, while, do, done, case, esac, function, return, local, export},
    morecomment=[l]{\#},
    morestring=[b]",
    morestring=[b]'
}

% ============================================================================
% METADATA
% ============================================================================
\title{Predicting Context Recovery Success in AI-Assisted Development}

% Authors (anonymized for review)
\author{Anonymous Author(s)}
\affiliation{
    \institution{Anonymous Institution}
    \city{City}
    \country{Country}
}
\email{anonymous@example.com}

% ============================================================================
% ABSTRACT
% ============================================================================
\begin{abstract}
AI-assisted software development has transformed how developers work, yet context loss during extended projects remains a critical challenge. When sessions span weeks or months, developers spend 15--25 minutes recovering context after each interruption. While workflow management systems can reduce this overhead, a fundamental question remains: Can we predict whether context recovery will succeed, and how long it will take?

We present the first predictive models for workflow context recovery in AI-assisted development. Using the Universal Workflow System (UWS) as our experimental platform, we generate a dataset of 3,000 recovery scenarios spanning diverse conditions: varying checkpoint counts (1--200), state complexities (minimal to complex), corruption levels (0--90\%), and interruption types (clean, crash, timeout). We train and evaluate machine learning models to predict recovery time, success probability, and state completeness.

Our results demonstrate that workflow recovery is predictable: Gradient Boosting achieves MAE of 1.1ms for recovery time prediction ($R^2 = 0.756$) and AUC-ROC of 0.912 for recovery success classification. Feature importance analysis reveals that corruption level ($r = -0.475$), checkpoint characteristics, and handoff document size are the dominant factors. We release a public benchmark dataset enabling future research on development workflow analytics.

\end{abstract}

% ============================================================================
% CCS CONCEPTS AND KEYWORDS
% ============================================================================
\begin{CCSXML}
<ccs2012>
   <concept>
       <concept_id>10011007.10011006.10011008.10011009.10011012</concept_id>
       <concept_desc>Software and its engineering~Software development techniques</concept_desc>
       <concept_significance>500</concept_significance>
   </concept>
   <concept>
       <concept_id>10010147.10010257.10010293.10010294</concept_id>
       <concept_desc>Computing methodologies~Supervised learning by regression</concept_desc>
       <concept_significance>500</concept_significance>
   </concept>
</ccs2012>
\end{CCSXML}

\ccsdesc[500]{Software and its engineering~Software development techniques}
\ccsdesc[500]{Computing methodologies~Supervised learning by regression}

\keywords{predictive models, workflow recovery, context management, machine learning, software analytics}

% ============================================================================
% DOCUMENT
% ============================================================================
\begin{document}

\maketitle

% Include PROMISE-specific sections
\input{sections/01-introduction-promise}
\input{sections/02-background}
\input{sections/03-approach}
\input{sections/04-evaluation-promise}
\input{sections/05-related}
\input{sections/06-conclusion}

% ============================================================================
% ACKNOWLEDGMENTS (hidden for review)
% ============================================================================
% \begin{acks}
% Acknowledgments here after acceptance.
% \end{acks}

% ============================================================================
% REFERENCES
% ============================================================================
\bibliographystyle{ACM-Reference-Format}
\bibliography{references}

% Balance columns on last page
\balance

\end{document}
